Very simplified likelihood

### 2023-11-28 inst/very_simplified_likelihood.Rmd
more work

Consider a race between $n+1$ competitors of strength $a,b,\ldots,b$
where we require the total strength to be unity: $a+nb=1$.  Suppose
our bro comes first:

\[\mathcal{L}(a)=a\]

Comes second:

\[
\mathcal{L}(a)=b\cdot\frac{a}{1-b}
\]

Third:

\[
\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{a}{1-2b}
\]

Fourth:

\[
\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{a}{1-3b}
\]

Placing after $r$ of the $b$ clones finish:

\[
\mathcal{L}(a)=\frac{b^ra}{(1-b)(1-2b)\cdots(1-rb)}=(1-nb)\frac{(-1)^r\Gamma(1-1/b)}{\Gamma(1+r-1/b)}
\]

```{r tryit}
small <- 1e-3

f_single_direct <- function(a,r,n,log=FALSE){  # not vectorised in anything
  b <- (1-a)/n   # 0 <= b <= 1/n
  if(log){
      return(log(a) + r*log(b) - sum(log(1-seq_len(r)*b)))
     } else {
      return(a * b^r/prod(1-seq_len(r)*b))
     }
}

`f_vectorised_direct` <- function(a,r,n,log=FALSE){  # vectorised in 'a', not in 'r'
  b <- (1-a)/n   # 0 <= b <= 1/n
  out <- r*log(b) + log(a) -rowSums(log(1-outer(b,seq_len(r))))
  if(log){return(out)} else {return(exp(out))}
}

f <- function(a,r,n,log=FALSE){
out <- a*0
if(log){
    for(rr in r){  out <- out + f_vectorised_direct(a,rr,n,log=TRUE)}
  } else {
    out[] <- 1
    for(rr in r){  out <- out * f_vectorised_direct(a,rr,n,log=FALSE)}
  }
  return(out)
}

f_gamma <- function(a,r,n,log=FALSE){ # vectorised in 'a', but screws up for b=1/integer
  b <- (1-a)/n   # 0 <= b <= 1/n
  out <- log(a) + lgamma(1-1/b) - lgamma(1+r-1/b)
  if(!log){out <- exp(out)}
  return(out)
}
```

Now use it:

```{r}
 a <- seq(from=0.05,to=0.45,by=0.01)
 L <- f(a,r=c(3,4,3,3,1,3,3,3,1,1,1,5),10,log=T)  # 10+1==11 competitors in total
plot(a,L-max(L))
abline(v=a[which.max(L)])
abline(v=1/11)
abline(h=c(0,-2))
```

Now skating:

```{r showhughes}
skating_table
hughes <- unlist(skating_table[1,])
a <- seq(from=0.05,to=0.55,by=0.01)
L <- f(a,hughes,23,log=TRUE)
L <- L - max(L)
plot(a,L,type='b')
abline(v=1/23)
```

```{r showall}
a <- seq(from=0.001,to=0.55,by=0.001)
fun <- function(i){
  out <- f(a,unlist(skating_table[i,]),23,log=TRUE)
  return(out-max(out))
}

plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
abline(v=1/23)
for(i in seq_len(9)){points(a,fun(i),type='b')}

plot(NA,xlim=c(-5,1),ylim=c(-5,0),type='n',xlab="log-odds")
abline(v=-log(22))
for(i in seq_len(23)){points(log(a/(1-a)),fun(i),type='l')}
```

# Formula 1

```{r showformula1}
x <- read.table("formula1_2022.txt",header=TRUE)
x <- x[,-ncol(x)]
x
```

```{r analyseform1}
fun <- function(i){
  suppressWarnings(jj <- as.numeric(unlist(x[i,])))
  jj[is.na(jj)] <- 23
  L <- f(a,jj,23,log=TRUE)
  return(L - max(L))
}

plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(22))
for(i in seq_len(20)){points(log(a/(1-a)),fun(i),type='l')}

```

# Mathematics Olympiad

```{r}
x <- as.matrix(read.table("olympiad.txt",header=TRUE))
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

```{r}
x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
```

```{r}
x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
x[] <- sample(c(x))
x
rowSums(x)
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

>>>>

Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  An
observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=n+(n-1)a$.

### Likelihood functions

Suppose our bro comes first (zero $b$'s ahead of him):

\[\mathcal{L}(a)=\frac{a}{a+nb}=\frac{a}{n+(n-1)a}\]

Comes second, one $b$ ahead of him:

\[
\mathcal{L}(a)=\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
\]

Third, two ahead of him:

\[
\mathcal{L}(a)=
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{a}{a+(n-2)b}
\]

Fourth, three ahead:

\[
\mathcal{L}(a)=
\frac{b}{a+ n   b}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{b}{a+(n-2)b}\cdot
\frac{a}{a+(n-3)b}
\]

Placing after $r$ of the $n$ clones of strength $b$ finish:

\[
\mathcal{L}(a)=
\frac{b^ra}{
(a+nb)(a+(n-1)b)(a+(n-2)b)\cdots(a+(n-r)b)}
=
\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}
\]

where $B=1/b$.  In R:

```{r}
f_single <- function(a,r,n,log=FALSE){  # not vectorised
  B <- 1/(1-a)
  if(log){
     out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
   } else {
     out <- (B-1)/prod(B+(n-r-1):(n-1))
   }
   return(out)
}

f_vec_a <- function(a,r,n, log=FALSE){  # vectorised in 'a' but not in 'r'
 sapply(a,function(a){f_single(a,r=r,n=n,log=log)})
}

f <- function(a,r,n,log=FALSE){ # vectorized in 'a', treats 'r' as a vector of independent observations

  if(log){
    out <- 0
    for(rr in r){out <- out + f_vec_a(a,rr,n,log=TRUE)}
  } else {
    out <- 1
    for(rr in r){out <- out * f_vec_a(a,rr,n,log=FALSE)}
  }
  return(out)
}

MLE <- function(r,n){
    d <- 1e-6
   optimize(f,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
}
MLE(3,9)
```

Now verification

```{r}
a <- 0.23423434
b <- 1-a
B <- 1/b
n <- 9
r <- 3
c(
b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
(B-1)/prod(B+(5:8)),
f(a,r,n,log=FALSE))
```

Now some visuals.  Suppose $r=4,n=8$:

```{r}
a <- seq(from=0,to=1,by=0.01)
plot(a,f(a,3,8,log=FALSE))
```

# Formula 1.

```{r maxvscheco}
a <- read.table("formula1_2023.txt",header=TRUE)
a <- a[,seq_len(ncol(a)-1)]
max <- as.numeric(as.matrix(a)[1,])
checo <- as.numeric(as.matrix(a)[2,])
checo[is.na(checo)] <- 22
max
checo
MLE(max,22)
MLE(checo,22)
```
